|

### 2021-08-11 prova_knit_finale.Rmd
bibliografia corretta
biblio
piccola modifica
cambia
nomi

author:
  - Edoardo Marcelli
  - Roberto Colarieti
  - Alexei Verkhovtsev
  - Michele Bolla
  - Gabriele Romano
  - Benedetta Bruni

@@

\subsection{Central limit theorem}
The last asymptotic result of PF estimators concerns convergence in law: the error of a particle estimate is asymptotically normally distributed. 
\bigskip
\begin{proposition}
For any $t\geq0$ and any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$
$$
\sqrt{N}\bigg( \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\E_{\Q_{t-1}M_t}(\varphi)\bigg)\implies \N(0,\hat{\mathcal{V}}_t(\varphi))
$$
and 
$$
\sqrt{N}\bigg(\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)-\E_{\Q_{t}}(\varphi)\bigg)\implies \N(0,\mathcal{V}_t(\varphi)),
$$
where the asymptotic variances are defined recursively, with $ \hat{\mathcal{V}}_0(\varphi):=Var_{\M_0}(\varphi) $,
\begin{equation*}
    \begin{split}
        \hat{\mathcal{V}}_t(\varphi)&:=\mathcal{V}_{t-1}(\E_{M_t}(\varphi))+Var_{\Q_{t-1}M_t}(\varphi) \text{ }\forall t\geq1, \\
        \mathcal{V}_t(\varphi)&:=\hat{\mathcal{V}}_{t}\big( \bar{G}_t(\varphi-\E_{\Q_t}(\varphi)) \big) \text{ }\forall t\geq0.
    \end{split}
\end{equation*}
Here, $\bar{G}_t=G_t/\E_{\Q_{t-1}M_t}(G_t)$ is the normalized incremental weight.
\end{proposition}
This version of the CLT for PF suggest that PF estimators with respect both to the predictive and the filtering distributions are approximately normal for a large number of particles. \
In particular, the logic behind the expression for $\mathcal{V}_t(\varphi)$ is the same as described in section \ref{pf_is_conv}. In fact, recall that the estimator with respect to the filtering distribution is a more sophisticated form of IS estimator, where the target distribution differs from the proposal one. \ 
Instead, to get an intuition of the variance $\hat{\mathcal{V}}_t(\varphi),$ note that by the law of total variance, 
\begin{equation*}
    \begin{split}
        Var\Bigg[ \frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n) \Bigg] = &Var\Bigg[ \E\bigg[\frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n)\Big| \sigma(X_{t-1}^{1:N}) \bigg] \Bigg] \\
        &+ \E \Bigg[ Var\bigg[ \frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n)    \Big| \sigma(X_{t-1}^{1:N})\bigg]\Bigg] \\
        &= Var\bigg[\sum\limits_{n=1}^N W_{t-1}^n (M_t\varphi)(X_{t-1}^n)\bigg] \\
        &+\frac{1}{N}\E\Bigg[Var\bigg[\varphi(X_t^1) \Big| \sigma(X_{t-1}^{1:N})\bigg]\Bigg].
    \end{split}
\end{equation*}
The notation $\mid \sigma(X_{t-1}^{1:N})$ means conditioning on having sampled (from the multinomial distribution) $X_{t-1}^{1:N}$. Conditional on multinomial sampling, the $X_{t-1}^n$ are i.i.d., with distribution  $\sum\limits_{n=1}^N W_{t-1}^n M_t(X_{t-1}^n, dx_t)$, that should converge to $\Q_{t-1}M_t$. So, it is natural to think that the second element of the sum converges to the variance of $\varphi$ with respect to the distribution $\Q_{t-1}M_t$. 
These results are of practical use because they provide a way to explicitly compute the variance of the PF estimates.

@@

\section{Bibliography}\

@@

\chapter*{Appendix}
\section{The Stochastic Volatility Model}

>>>>

author:
  - Michele Bolla
  - Benedetta Bruni
  - Roberto Colarieti
  - Edoardo Marcelli
  - Gabriele Romano
  - Alexei Verkhovtsev

@@

\subsection{Central limit theorem}
The last asymptotic result of PF estimators concerns convergence in law: the error of a particle estimate is asymptotically normally distributed. 
\bigskip
\begin{proposition}
For any $t\geq0$ and any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$
$$
\sqrt{N}\bigg( \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\E_{\Q_{t-1}M_t}(\varphi)\bigg)\implies \N(0,\hat{\mathcal{V}}_t(\varphi))
$$
and 
$$
\sqrt{N}\bigg(\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)-\E_{\Q_{t}}(\varphi)\bigg)\implies \N(0,\mathcal{V}_t(\varphi)),
$$
where the asymptotic variances are defined recursively, with $ \hat{\mathcal{V}}_0(\varphi):=Var_{\M_0}(\varphi) $,
\begin{equation*}
    \begin{split}
        \hat{\mathcal{V}}_t(\varphi)&:=\mathcal{V}_{t-1}(\E_{M_t}(\varphi))+Var_{\Q_{t-1}M_t}(\varphi) \text{ }\forall t\geq1, \\
        \mathcal{V}_t(\varphi)&:=\hat{\mathcal{V}}_{t}\big( \bar{G}_t(\varphi-\E_{\Q_t}(\varphi)) \big) \text{ }\forall t\geq0.
    \end{split}
\end{equation*}
Here, $\bar{G}_t=G_t/\E_{\Q_{t-1}M_t}(G_t)$ is the normalized incremental weight.
\end{proposition}
This version of the CLT for PF suggest that PF estimators with respect both to the predictive and the filtering distributions are approximately normal for a large number of particles. \
In particular, the logic behind the expression for $\mathcal{V}_t(\varphi)$ is the same as described in section \ref{pf_is_conv}. In fact, recall that the estimator with respect to the filtering distribution is a more sophisticated form of IS estimator, where the target distribution differs from the proposal one. \ 
Instead, to get an intuition of the variance $\hat{\mathcal{V}}_t(\varphi),$ note that by the law of total variance, 
\begin{equation*}
    \begin{split}
        Var\Bigg[ \frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n) \Bigg] = &Var\Bigg[ \E\bigg[\frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n)\Big| \sigma(X_{t-1}^{1:N}) \bigg] \Bigg] \\
        &+ \E \Bigg[ Var\bigg[ \frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n)    \Big| \sigma(X_{t-1}^{1:N})\bigg]\Bigg] \\
        &= Var\bigg[\sum\limits_{n=1}^N W_{t-1}^n (M_t\varphi)(X_{t-1}^n)\bigg] \\
        &+\frac{1}{N}\E\Bigg[Var\bigg[\varphi(X_t^1) \Big| \sigma(X_{t-1}^{1:N})\bigg]\Bigg].
    \end{split}
\end{equation*}
The notation $\mid \sigma(X_{t-1}^{1:N})$ means conditioning on having sampled (from the multinomial distribution) $X_{t-1}^{1:N}$. Conditional on multinomial sampling, the $X_{t-1}^n$ are i.i.d., with distribution  $\sum\limits_{n=1}^N W_{t-1}^n M_t(X_{t-1}^n, dx_t)$, that should converge to $\Q_{t-1}M_t$. So, it is natural to think that the second element of the sum converges to the variance of $\varphi$ with respect to the distribution $\Q_{t-1}M_t$. 
These results are of practical use because they provide a way to explicitly compute the asymptotic variance of the PF estimates.

@@

\chapter*{Bibliography}\

@@

\chapter*{Appendix}
\section*{The Stochastic Volatility Model}
